---
title: "HTO mouse T-cell dataset — full analysis from raw data (scroll)"
subtitle: "Cao et al. — raw 10x h5 → gated singlets → cell types → subset views → scroll explorer"
author: "Generated for verification"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    df_print: paged
    fig_width: 7
    fig_height: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  cache = TRUE, cache.lazy = FALSE, fig.align = "center"
)
set.seed(1)
```

This document reproduces the **entire pipeline from raw data** for the two-sample
mouse RNA + hashtag (HTO) experiment: from the raw 10x `.h5` matrices through
gated singlets, cell-type annotation, per-cell-type **subset views**, and the
final `scroll` explorer with a counts store for pseudobulk DE. Knit it to verify
every intermediate output.

The only non-scriptable step is the **manual hashtag gating** (interactive), which
was performed once in scroll's gating app and exported to a CSV; this document
consumes that export and runs everything else from the raw matrices.

Heavy chunks are cached, so the first knit is slow (~20–30 min: several
re-embeddings of tens of thousands of cells) and re-knits are fast.

```{r libs}
suppressPackageStartupMessages({
  library(Seurat); library(SeuratObject); library(Matrix)
  library(ggplot2); library(dplyr)
  library(scroll)     # the package under test
})

root   <- "/Users/shihanl1/scroll"
h5s    <- c(s1 = file.path(root, "data/s1_sample_filtered_feature_bc_matrix.h5"),
            s2 = file.path(root, "data/s2_sample_filtered_feature_bc_matrix.h5"))
gating <- file.path(root, "data/hto_keep_barcodes.csv")   # export of the manual gating
exdir  <- file.path(root, "examples/hto-filter")
outdir <- file.path(exdir, "project-filtered")            # scroll project (git-ignored)
stopifnot(all(file.exists(h5s)), file.exists(gating))

MT_MAX    <- 10     # percent.mt filter
NFEAT_MAX <- 6000   # nFeature_RNA filter
RES       <- seq(0.1, 1.0, by = 0.1)
```

# 1. Raw 10x → merged RNA + HTO object

Each `.h5` holds **Gene Expression** (RNA) and **Antibody Capture** (the three
hashtags). Build one Seurat object per sample with an `HTO` assay, prefix
barcodes by sample so ids stay unique, merge, and compute mouse mito %. (Seurat
sanitizes the hashtag underscores to dashes, so features become `Hashtag-1/2/3`.)

```{r read-raw}
make_obj <- function(h5, sample) {
  x   <- Read10X_h5(h5)
  obj <- CreateSeuratObject(counts = x[["Gene Expression"]], project = sample)
  obj[["HTO"]] <- CreateAssay5Object(counts = x[["Antibody Capture"]])
  obj$sample <- sample
  obj
}
o1 <- make_obj(h5s[["s1"]], "s1")
o2 <- make_obj(h5s[["s2"]], "s2")

obj <- merge(o1, o2, add.cell.ids = c("s1", "s2"))
obj <- JoinLayers(obj); obj <- JoinLayers(obj, assay = "HTO")
obj[["percent.mt"]] <- PercentageFeatureSet(obj, pattern = "^mt-")   # mouse mito

data.frame(sample = c("s1", "s2", "merged"),
           cells  = c(ncol(o1), ncol(o2), ncol(obj)),
           genes  = c(nrow(o1), nrow(o2), nrow(obj)))
```

## HTO CLR normalization

`margin = 2` normalizes each hashtag across cells, so an independent threshold per
hashtag separates positive from background. The per-cell CLR values are copied
into metadata (`hto_Hashtag-*`) — these are what the gating and the Biaxial panel
plot.

```{r hto-clr}
obj <- NormalizeData(obj, assay = "HTO", normalization.method = "CLR",
                     margin = 2, verbose = FALSE)
hto_clr <- as.matrix(GetAssayData(obj, assay = "HTO", layer = "data"))   # 3 x cells
for (h in rownames(hto_clr)) obj[[paste0("hto_", h)]] <- hto_clr[h, ]
obj <- SetAssayData(obj, assay = "HTO", layer = "data",
                    new.data = Matrix::Matrix(hto_clr, sparse = TRUE))
rownames(hto_clr)
```

# 2. Hashtag demultiplexing (manual gating → CSV)

The three hashtags have tricky background, so single-positive gates were drawn
**manually** in scroll's QC/hashtag gating panel (`examples/hto-filter/panel.R`)
and exported to `data/hto_keep_barcodes.csv` (barcode + assignment). That step is
interactive and shown here for provenance only:

```{r gating-provenance, eval=FALSE}
# NOT run — interactive. Reproduce with:
#   scroll::scroll_serve("examples/hto-filter/project")  # built by build.R
#   -> QC & hashtag gating section -> draw single-positive gates
#   -> Download CSV -> data/hto_keep_barcodes.csv
```

We load the export and attach the assignment (Singlet hashtag / Doublet /
Negative) as metadata.

```{r gating-load}
keep <- utils::read.csv(gating, stringsAsFactors = FALSE)
obj$assignment <- keep$assignment[match(colnames(obj), keep$cell)]
table(obj$assignment, useNA = "ifany")
```

The demux is exactly what the new **Biaxial** panel shows — pairwise hashtag CLR
scatters coloured by the assignment. Reproduced here with the package's own view:

```{r gating-biaxial, fig.width=10, fig.height=3.6}
hto_cols <- grep("^hto_", colnames(obj[[]]), value = TRUE)
gate_df <- obj[[]]
gate_df$assignment[is.na(gate_df$assignment)] <- "Unassigned"
view_biaxial(gate_df, list(features = hto_cols, color_by = "assignment"),
             list(point_size = 0.3, alpha = 0.5))
```

# 3. Singlets + QC filtering + re-embedding

Keep **singlets** (assignment is one hashtag) that also pass `percent.mt < 10` and
`nFeature_RNA < 6000`, then re-process RNA from counts on the clean population.

```{r reprocess}
obj$hto <- ifelse(grepl("^Hashtag", obj$assignment), obj$assignment, NA_character_)
keep_cells <- colnames(obj)[
  !is.na(obj$hto) & obj$percent.mt < MT_MAX & obj$nFeature_RNA < NFEAT_MAX]
sub <- subset(obj, cells = keep_cells)

DefaultAssay(sub) <- "RNA"
sub <- sub |>
  NormalizeData(verbose = FALSE) |>
  FindVariableFeatures(verbose = FALSE) |>
  ScaleData(verbose = FALSE) |>
  RunPCA(npcs = 30, verbose = FALSE) |>
  RunUMAP(dims = 1:30, verbose = FALSE)

data.frame(stage = c("merged", "singlets", "+ QC filters (kept)"),
           cells = c(ncol(obj), sum(!is.na(obj$hto)), ncol(sub)))
```

```{r qc-plots, fig.width=9, fig.height=3.2}
VlnPlot(sub, c("nCount_RNA", "nFeature_RNA", "percent.mt"),
        group.by = "sample", pt.size = 0, ncol = 3) & theme(legend.position = "none")
```

# 4. Genotype & treatment from hashtag / sample

The 3 hashtags encode the CRISPR guide (genotype); the 2 samples encode
treatment. Derived deterministically and cross-checked.

```{r meta}
geno_map <- c("Hashtag-1" = "sgCtrl", "Hashtag-2" = "sgIcos", "Hashtag-3" = "sgCxcr5")
trt_map  <- c("s1" = "post", "s2" = "neg")
sub$genotype  <- factor(unname(geno_map[sub$hto]),   levels = c("sgCtrl","sgIcos","sgCxcr5"))
sub$treatment <- factor(unname(trt_map[sub$sample]), levels = c("post","neg"))
list(genotype = table(sub$genotype), treatment = table(sub$treatment),
     `hashtag x genotype` = table(sub$hto, sub$genotype))
```

# 5. Clustering (res 0.1), cell-type labels, and marker ID

```{r cluster01}
sub <- FindNeighbors(sub, dims = 1:30, verbose = FALSE)
sub <- FindClusters(sub, resolution = 0.1, verbose = FALSE)
sub$cluster01 <- sub$seurat_clusters
table(sub$cluster01)
```

```{r labels}
label_map <- c("0"="Cd101+","1"="Tfh","2"="Cx3cr1+","3"="Cycling",
               "4"="C4","5"="C5","6"="B cells")
sub$celltype <- factor(unname(label_map[as.character(sub$cluster01)]),
                       levels = c("Cd101+","Tfh","Cx3cr1+","Cycling","C4","C5","B cells"))
table(sub$celltype)
```

```{r umap-labeled, fig.width=7.5, fig.height=5}
DimPlot(sub, group.by = "celltype", label = TRUE, repel = TRUE) +
  ggtitle("res-0.1 cell types")
```

## Marker identification (one-vs-rest, presto)

```{r markers}
mk <- presto::wilcoxauc(GetAssayData(sub, assay = "RNA", layer = "data"),
                        as.character(sub$celltype))
top_markers <- function(group, n = 12) {
  g <- mk[mk$group == group, ]
  head(g[order(-g$auc), c("feature","auc","logFC","pct_in","pct_out")], n)
}
setNames(lapply(levels(sub$celltype)[1:6], top_markers), levels(sub$celltype)[1:6])
```

**Reading of the two unnamed clusters (hypotheses):**

- **C4** — TCR-signaling genes (`Tox`, `Themis`, `Fyn`, `Dock2/10`, `Ikzf1`,
  `Grap2`) over a high-lncRNA background (`Malat1`, `Gm42418`): a **resting /
  naive-like or Tox-hi transitional T** state rather than a distinct effector
  program. Check naive vs exhaustion markers below.
- **C5** — dominated by **mitochondrial** genes (`mt-Co1/2/3`, `mt-Atp8`,
  `mt-Cytb`) with `Malat1`, still `Ptprc+`: a **high-mito / stressed / low-quality**
  signature, likely partly technical.

```{r c45-diagnostics, fig.width=9, fig.height=5.5}
FeaturePlot(sub, c("Tcf7","Sell","Pdcd1","Tox","Ptprc","percent.mt"),
            order = TRUE, ncol = 3)
```

# 6. Remove B cells, re-embed the main object (merge C4/5)

B cells (cluster 6) are removed and the rest re-embedded from scratch — the app's
**main UMAP**. Clusters 4 and 5 are merged into one **C4/5** cell type. A global
resolution sweep 0.1–1.0 is computed on the new graph.

```{r main-reembed}
main <- subset(sub, cells = colnames(sub)[sub$celltype != "B cells"])
ct <- as.character(main$celltype); ct[ct %in% c("C4","C5")] <- "C4/5"
main$celltype <- factor(ct, levels = c("Cd101+","Tfh","Cx3cr1+","Cycling","C4/5"))

DefaultAssay(main) <- "RNA"
main <- main |>
  NormalizeData(verbose = FALSE) |> FindVariableFeatures(verbose = FALSE) |>
  ScaleData(verbose = FALSE) |> RunPCA(npcs = 30, verbose = FALSE) |>
  RunUMAP(dims = 1:30, verbose = FALSE) |> FindNeighbors(dims = 1:30, verbose = FALSE)

res_names <- sprintf("res_%.1f", RES)
for (i in seq_along(RES)) {
  main <- FindClusters(main, resolution = RES[i], verbose = FALSE)
  main[[res_names[i]]] <- paste0("c", as.character(main$seurat_clusters))
}
main[["seurat_clusters"]] <- NULL
c(cells = ncol(main),
  setNames(vapply(res_names, function(n) length(unique(main[[n]][[1]])), 0L), res_names))
```

```{r main-umaps, fig.width=10, fig.height=3.4}
patchwork::wrap_plots(
  DimPlot(main, group.by = "celltype",  label = TRUE, repel = TRUE) + ggtitle("cell type"),
  DimPlot(main, group.by = "genotype")  + ggtitle("genotype"),
  DimPlot(main, group.by = "treatment") + ggtitle("treatment"),
  nrow = 1)
```

# 7. Per-cell-type subset views (re-embed + subcluster sweep)

Each of the 5 cell types is re-embedded **in isolation** and re-clustered across
resolutions 0.1–1.0. These become the app's **subset views** (attached by barcode
with `scroll_add_subset()`; subcluster columns are scoped to their view).

```{r subsets}
ids <- c(cd101="Cd101+", tfh="Tfh", cx3cr1="Cx3cr1+", cycling="Cycling", c45="C4/5")
sub_meta <- list(); sub_umaps <- list()

for (id in names(ids)) {
  lab <- unname(ids[id])
  s <- subset(main, cells = colnames(main)[main$celltype == lab])
  DefaultAssay(s) <- "RNA"
  s <- s |>
    NormalizeData(verbose = FALSE) |> FindVariableFeatures(verbose = FALSE) |>
    ScaleData(verbose = FALSE) |> RunPCA(npcs = 30, verbose = FALSE) |>
    RunUMAP(dims = 1:30, verbose = FALSE) |> FindNeighbors(dims = 1:30, verbose = FALSE)

  cols <- sprintf("%s_res_%.1f", id, RES)
  for (i in seq_along(RES)) {
    s <- FindClusters(s, resolution = RES[i], verbose = FALSE)
    v <- rep(NA_character_, ncol(main)); names(v) <- colnames(main)
    v[colnames(s)] <- paste0("c", as.character(s$seurat_clusters))
    main[[cols[i]]] <- unname(v)
  }
  main <- scroll_add_subset(main, s, name = id, label = lab,
                            embeddings = setNames("umap", paste0("umap_", id)), meta = cols)
  sub_meta[[id]]  <- cols
  sub_umaps[[id]] <- DimPlot(s, group.by = "RNA_snn_res.0.3") +
    ggtitle(sprintf("%s (%d cells)", lab, ncol(s))) + theme(legend.position = "none")
}
```

```{r subset-umap-grid, fig.width=10, fig.height=6.5}
patchwork::wrap_plots(sub_umaps, ncol = 3)
```

# 8. Composition by genotype (biology check)

ICOS and CXCR5 are central to the Tfh program, so guide knockouts are expected to
shift composition — especially the **Tfh** fraction.

```{r composition, fig.width=7.5, fig.height=4.5}
comp <- main[[]] |>
  dplyr::count(genotype, celltype) |>
  dplyr::group_by(genotype) |> dplyr::mutate(frac = n / sum(n)) |> dplyr::ungroup()
ggplot(comp, aes(genotype, frac, fill = celltype)) +
  geom_col() + scale_y_continuous(labels = scales::percent) +
  labs(y = "fraction of cells", x = NULL, title = "Cell-type composition by genotype") +
  theme_minimal(base_size = 12)
```

```{r composition-table}
tidyr::pivot_wider(comp[, c("genotype","celltype","frac")],
                   names_from = celltype, values_from = frac) |>
  dplyr::mutate(dplyr::across(where(is.numeric), ~round(.x, 3)))
```

# 9. Build the scroll project (subset views + counts store)

Persist the object and build the explorer with `counts = TRUE` (enables Pseudobulk
DE). Brand it "Cao et al." (the config regenerates on each build).

```{r build, cache=FALSE}
main <- SetAssayData(main, assay = "HTO", layer = "data",
  new.data = as(GetAssayData(main, assay = "HTO", layer = "data"), "CsparseMatrix"))
saveRDS(main, file.path(exdir, "filtered-subsets.rds"))

subsets <- lapply(names(ids), function(id)
  list(label = unname(ids[id]), embeddings = paste0("umap_", id), meta = sub_meta[[id]]))
names(subsets) <- names(ids)
hto_cols <- grep("^hto_", colnames(main[[]]), value = TRUE)

scroll_build(
  main, outdir, assays = c("RNA", "HTO"), counts = TRUE,
  meta_cols = c("sample","hto","genotype","treatment",
                "nCount_RNA","nFeature_RNA","percent.mt","celltype",
                res_names, hto_cols, unlist(sub_meta, use.names = FALSE)),
  subsets = subsets, overwrite = TRUE)

cfg <- yaml::read_yaml(file.path(outdir, "config.yaml"))
cfg$brand <- "Cao et al."; cfg$title <- "Mouse Tfh CRISPR (HTO)"
yaml::write_yaml(cfg, file.path(outdir, "config.yaml"))

m <- scroll_manifest(outdir)
list(n_cells = m$n_cells, has_counts = m$has_counts,
     subsets = names(m$subsets), embeddings = names(m$embeddings))
```

# 10. Pseudobulk DE example — sgIcos vs sgCtrl within Tfh

A worked verification of the Pseudobulk DE panel. There are **no biological
replicates** (one hashtag per genotype), so this uses `no_replicate`
pseudo-replication — anti-conservative by construction; treat p-values as a
ranking, not inference. `Icos` itself should drop in the sgIcos arm.

```{r pseudobulk, cache=FALSE, fig.width=6.5, fig.height=5}
data <- scroll:::.scroll_load(outdir)
tfh <- data$cells[data$cells$celltype == "Tfh", ]
res <- scroll_pseudobulk_de(
  data, assay = "RNA", aggregate_cols = "genotype",
  ident1 = "sgIcos", ident2 = "sgCtrl", replicate_col = "no_replicate",
  n_pseudo = 4, cells_per_pseudo = 200, min_cells = 50, cells = tfh)

knitr::kable(head(res[order(res$p_val_adj), ], 15), digits = 3,
             caption = "Top pseudobulk DE genes (sgIcos vs sgCtrl, Tfh)")
print(view_volcano(res, list(lfc = 1, padj = 0.05, labeln = 15)))
cat("Icos logFC (expect negative in sgIcos):",
    round(res$logFC[res$gene == "Icos"], 3), "\n")
scroll_disconnect(data$con)
```

# Session info

```{r sessioninfo, cache=FALSE}
sessionInfo()
```
