---
title: "Getting started: from a Seurat object to a running app"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Getting started: from a Seurat object to a running app}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE          # every chunk is narrated: it needs SeuratData, writes
)                       # files, or launches a server, so none run at build time
```

scroll turns a processed Seurat object with clustering, an embedding,
and normalized expression into an extendable interactive Shiny app. The
objective of this article is to introduce the basic workflow of the app
package using the classic 3K PBMC dataset.

## 1. Introduction

scroll works in two separate phases. First, the build call reads the
object and writes small, flat files to disk. The light *runtime* app
reads only those files.

With a pre-processed single-cell object, an app can be created through
two function calls.

```{r}
library(scroll)

# Phase 1 — offline, run once: extract flat artifacts from the object
scroll_build(obj, "pbmc3k")

# Phase 2 — the running app: reads only those artifacts
scroll_app("pbmc3k")
```

To preserve memory, count tables are converted into parquet tables which
are accessed in-time. While this significantly reduces RAM usage after
deployment, the build times can be quite slow, specifically for larger
Seurat objects.

### What you'll need

-   R 4.1 or newer. scroll is an R package; you will work from an R
    console throughout.
-   A processed Seurat object

## 2. Install scroll and its dependencies

scroll is installed from GitHub using `remotes`. Its required
dependencies are all on CRAN and are pulled in automatically.

```{r}
install.packages("remotes")
remotes::install_github("shihanli1992/scroll")
```

### Optional features

Several capabilities are optional. Install only what you need.

| Package(s)               | Unlocks                                              |
|--------------------------|------------------------------------------------------|
| `presto`                 | Live Wilcoxon differential expression (the DE panel) |
| `edgeR`, `limma`         | Replicate-aware pseudobulk DE (Bioconductor)         |
| `ggtree`, `aplot`, `ape` | Hierarchical-clustering dendrograms on the DotPlot   |
| `scattermore`            | Rasterized scatter                                   |
| `shinycssloaders`        | Loading spinners while plots render                  |
| `DT`, `ggrepel`          | Interactive DE tables and gene labels on the volcano |
| `Signac`                 | scATAC-seq (peaks) support                           |

```{r}
# Bioconductor packages (edgeR/limma -> pseudobulk DE; ggtree -> dendrograms; Signac -> ATAC)
install.packages("BiocManager")
BiocManager::install(c("edgeR", "limma", "ggtree", "Signac"))

# CRAN packages
install.packages(c("aplot", "ape", "scattermore", "shinycssloaders", "DT", "ggrepel"))

# presto — live Wilcoxon DE (from GitHub)
remotes::install_github("immunogenomics/presto")
```

### Packages for this tutorial

The worked example in the next steps builds from a small public dataset,
which needs Seurat's object class and the example-data package:

```{r}
install.packages("SeuratObject")
remotes::install_github("satijalab/seurat-data")
```

> Only for the build phase. `SeuratObject` is needed to *read* your
> object during `scroll_build()`. It is not required to run a built app,
> so a deployment server never needs Seurat installed.

Finally, load scroll and confirm the version:

```{r}
library(scroll)
packageVersion("scroll")
```

```         
#> [1] '0.2.0'
```

## 3. Loading an example object

```{r}
library(SeuratData)

InstallData("pbmc3k")
pbmc <- LoadData("pbmc3k", type = "pbmc3k.final")
pbmc <- UpdateSeuratObject(pbmc)
pbmc
```

```         
#> An object of class Seurat
#> 13714 features across 2638 samples within 1 assay
#> Active assay: RNA (13714 features, 2000 variable features)
#>  3 layers present: counts, data, scale.data
#>  2 dimensional reductions calculated: pca, umap
```

That print-out already tells you the three things scroll cares about: an
assay with a `data` layer, and two reductions (`pca` and `umap`).

For scroll to run, it requires a "processed" object containing all three
of the following:

-   At least one dimensional reduction (an embedding).
-   A normalized `data` layer in the assay.
-   Cell metadata.

```{r}
Reductions(pbmc)          # embeddings scroll can plot
```

```         
#> [1] "pca"  "umap"
```

```{r}
Layers(pbmc[["RNA"]])     # need a normalized "data" layer, not just counts
```

```         
#> [1] "counts"     "data"       "scale.data"
```

```{r}
colnames(pbmc[[]])        # the metadata columns
```

```         
#> [1] "orig.ident"         "nCount_RNA"         "nFeature_RNA"
#> [4] "seurat_annotations" "RNA_snn_res.0.5"    "seurat_clusters"
```

The exact metadata columns vary by object; in the pbmc3k dataset,
`seurat_annotations` holds the cell-type labels. You don't need to
curate this list now: in the next step scroll infers sensible reductions
and metadata automatically, and step 7 shows how to choose them
explicitly.

> Multiple assays and reductions are fine. If your object carries
> several assays (e.g. an ADT or ATAC assay) or extra reductions, these
> are automatically or manually added to the app.

## 4. Build a project with `scroll_build()`

`scroll_build()` reads the object and writes a self-contained project
directory. Without additional inputs, the function uses the object's
default assay, includes every reduction, and infers which metadata
columns to expose.

```{r}
library(scroll)

scroll_build(pbmc, "pbmc3k")
```

### What got written

```{r}
list.files("pbmc3k")
```

```         
#> [1] "app.R"         "cells.parquet" "config.yaml"   "expr"          "manifest.yaml"
```

| Artifact        | What it is                                                                                                         |
|-----------------|--------------------------------------------------------------------------------------------------------------------|
| `cells.parquet` | One row per cell containing metadata plus embedding coordinates.                                                   |
| `expr/`         | Expression as a sparse, zstd-compressed Parquet store, bucketed by the first letter of each gene.                  |
| `manifest.yaml` | A machine-readable description of the build including the assays, features, embeddings, metadata and their levels. |
| `config.yaml`   | Hand-editable app settings                                                                                         |
| `app.R`         | The ready-to-run entry point: `library(scroll); scroll_app(".")`.                                                  |

> Rebuilding. If the output directory already exists and isn't empty,
> `scroll_build()` will stop by default. Pass `overwrite = TRUE` to
> rebuild.

### Inspect the manifest

You rarely need to open the manifest by hand, but it's a good way to
confirm the build captured what you expected. `scroll_manifest()` reads
it back as a plain list.

```{r}
m <- scroll_manifest("pbmc3k")
m$n_cells
names(m$embeddings)
```

```         
#> [1] 2638
#> [1] "pca"  "umap"
```

## 5. Running the app

`scroll_serve()` starts the app locally and opens it in your browser.

```{r}
scroll_serve("pbmc3k")                 # launches locally, opens a browser
scroll_serve("pbmc3k", port = 8080)    # extra args pass through to shiny::runApp()
```

The project's own `app.R` does the same thing
(`library(scroll); scroll_app(".")`), so you can also open
`pbmc3k/app.R` in RStudio and click Run App, or run
`shiny::runApp("pbmc3k")`.

The window has three parts:

-   An app bar across the top displays the dataset title, a live cell
    count, and a global subset filter that narrows *every* panel at
    once.
-   A panel rail down the side for quick access to each panel.
-   The active panel with its own controls on the left, the plot on the
    right.

### The built-in panels

The following panels are pre-built and generated by default:

-   DimPlot: the embedding coloured by a metadata column.
-   FeaturePlot: the embedding coloured by a gene's expression.
-   Biaxial: a scatter of one numeric axis against another: two QC
    metrics, or two genes.
-   DotPlot: marker genes by group and dot size.
-   Violin: the distribution of a gene's expression across groups.
-   Proportions: stacked composition bars of one categorical column
    within another.
-   DE: live differential expression between groups (Wilcoxon, via
    `presto`), shown as a sortable table and a volcano plot.

> Panels that require uninstalled packages will show a message instead
> of a plot. For example, the DE panel needs `presto`. Without it the
> panel loads but reports that live DE is unavailable.

## 6. Customize the app with `config.yaml`

The app auto-populates every control from the manifest, so it works with
no configuration at all. `config.yaml` only sets *defaults* and labels.
Because it's read at runtime, editing it needs no rebuild: change the
file, restart the app, done. The expression store is never touched.

Here's the starter file scroll wrote for `pbmc3k`:

``` yaml
default_assay: RNA
default_embedding: umap
markers:
- CD3D
- CD8A
- IL7R
- CCR7
- MS4A1
- CD79A
- CD14
- LYZ
```

scroll picked those marker genes automatically from a list of common
ones present in the data. Note there's no `title` --- that's the first
thing worth adding.

### Title, defaults, and markers

Open `pbmc3k/config.yaml` in any text editor and edit it.

For example, add a title and curate the DotPlot's default genes:

``` yaml
title: PBMC 3k                # labels the app bar (add this)
default_assay: RNA
default_embedding: umap       # which reduction opens first
markers:                      # the DotPlot's default gene set
- MS4A1
- CD79A
- CD3D
- CD8A
- GNLY
- NKG7
- CD14
- LYZ
```

| Key                 | Effect                                                                                          |
|---------------------|-------------------------------------------------------------------------------------------------|
| `title`             | Text shown in the app bar. Omit it and the bar shows just the brand.                            |
| `default_embedding` | The reduction DimPlot / FeaturePlot open on (any name from the manifest, e.g. `umap` or `pca`). |
| `default_assay`     | The assay selected first when the object has more than one.                                     |
| `markers`           | The genes pre-loaded into the DotPlot. Any gene present in the assay.                           |

Restart the app to see the changes:

```{r}
scroll_serve("pbmc3k")   # the app bar now reads "PBMC 3k"; DotPlot opens on your genes
```

### Which panels appear

By default the panels are chosen automatically. Two optional keys let
you take over. `panels:` lists the exact panels to show, in this order,
overriding the automatic gating entirely. `exclude_panels:` instead
keeps the automatic set but hides a few.

``` yaml
# show exactly these, in this order:
panels:
- dimplot
- featureplot
- dotplot
- de

# — or — keep the automatic set but hide some:
exclude_panels:
- biaxial
```

Panel ids are the lowercase names: `dimplot`, `featureplot`, `biaxial`,
`dotplot`, `violin`, `proportions`, `de`, `pseudobulk`. You'd typically
use one key or the other, not both.

## 7. Customising the build

### Metadata, assays, and embeddings

Left to infer, scroll will select all columns with reasonable
cardinality. Specific columns can also be manually selected.

```{r}
scroll_build(pbmc, "pbmc3k",
             assays     = "RNA",
             embeddings = c("umap", "pca"),
             meta_cols  = c("seurat_annotations", "seurat_clusters",
                            "nCount_RNA", "nFeature_RNA"),
             overwrite  = TRUE)
```

| Argument     | What it controls                                                            |
|--------------|-----------------------------------------------------------------------------|
| `meta_cols`  | Which metadata columns become app selectors.                                |
| `assays`     | Which assays to include.                                                    |
| `embeddings` | Which reductions to carry. Any 2-D reduction becomes a plottable embedding. |

### `quantize`: size vs. exactness

By default scroll compresses expression values to 8-bit integers about
4× smaller on disk,. The cost is that very small values (below roughly
`max / 510`) floor to zero, so percent-expressing and dot sizes slightly
under-count the lowest expressors. Pass `quantize = FALSE` for exact
`float32` values.

| Setting            | Store        | Precision             | Use when                               |
|--------------------|--------------|-----------------------|----------------------------------------|
| `quantize = TRUE`  | \~4× smaller | low values floor to 0 | almost always --- the default          |
| `quantize = FALSE` | larger       | exact                 | you need precise low-expression values |

### `counts`

Remember the Pseudobulk DE panel that was missing in step 5? It needs
raw counts, which the default build doesn't store. Add `counts = TRUE`
and rebuild:

```{r}
scroll_build(pbmc, "pbmc3k", counts = TRUE, overwrite = TRUE)
scroll_serve("pbmc3k")   # the Pseudobulk DE panel now appears
```

This writes an extra, unquantized counts store (so the project roughly
doubles in size) and flips `has_counts` in the manifest, which is what
the panel's gate checks. It also needs `edgeR` and `limma` from step 2.

## 8. Subset views

Subsetted cells with their own embeddings and clusters can be added to
the app.

### Reprocess a slice, then attach it

Subset T cells and give them their own embedding:

```{r}
library(Seurat)

tcells <- subset(pbmc, subset = seurat_annotations %in%
                   c("Naive CD4 T", "Memory CD4 T", "CD8 T"))

tcells <- NormalizeData(tcells)      |>
  FindVariableFeatures()             |>
  ScaleData()                        |>
  RunPCA()                           |>
  RunUMAP(dims = 1:20)               # a fresh UMAP for just these cells

tcells <- FindNeighbors(tcells, dims = 1:20) |> FindClusters(resolution = 0.4)
tcells$tcell_subcluster <- tcells$seurat_clusters
```

`scroll_add_subset()` copies the child's reduction and chosen metadata
onto the parent, matched by cell barcode (cells outside the slice get
`NA`), and records the view so the build finds it automatically.

```{r}
# attach the sub-embedding + sub-labels onto the parent, matched by barcode
pbmc <- scroll_add_subset(pbmc, tcells,
                          name       = "tcell",
                          label      = "T cells",
                          embeddings = c(umap_tcell = "umap"),   # child "umap" -> parent "umap_tcell"
                          meta       = "tcell_subcluster")

# rebuild — the recorded subset is picked up automatically
scroll_build(pbmc, "pbmc3k", counts = TRUE, overwrite = TRUE)
```

The `embeddings` argument is a named vector,
`new_name = source_reduction`: it takes the reduction called `umap` in
`tcells` and stores it on the parent as `umap_tcell`. (If you prefer,
you can skip the helper and pass a `subsets =` list straight to
`scroll_build()`.)
