---
title: "Getting started: from a Seurat object to a running app"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started: from a Seurat object to a running app}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`scroll` turns a processed Seurat object into a polished, interactive single-cell
explorer. It has **two phases**:

1. A heavy, one-time **build** (`scroll_build()`) that extracts lightweight
   on-disk artifacts — cell metadata + embeddings as one small Parquet table, and
   expression as a feature-partitioned Parquet store.
2. A light **runtime** (`scroll_app()` / `scroll_serve()`) — a bslib Shiny app
   that reads only those artifacts, querying expression **one gene at a time**
   through arrow. The running app never loads the Seurat object, so runtime
   memory stays flat regardless of dataset size, and it deploys as a plain `app.R`.

The code chunks below are illustrative (they need a dataset / launch a server), so
they are not executed when this vignette is rendered.

## Install

```{r install}
install.packages(c("shiny", "bslib", "arrow", "dplyr", "ggplot2",
                   "scales", "yaml", "Matrix", "colourpicker"))
install.packages("SeuratObject")            # the build step needs this (runtime does not)
# then install scroll itself: R CMD INSTALL scroll  (or install.packages the tarball)
```

## 1. Build a project

Start from any processed Seurat object with at least one dimensional reduction and
a non-empty `data` layer. Here we use the `pbmc3k` demo from **SeuratData**:

```{r build}
library(scroll)
library(SeuratData)                    # for the example object
data("pbmc3k.final", package = "pbmc3k.SeuratData")
obj <- SeuratObject::UpdateSeuratObject(pbmc3k.final)   # ensure a v5 object

scroll_build(obj, "pbmc3k")            # object -> project directory
```

`scroll_build()` reads the object once and writes a self-contained project:

```
pbmc3k/
  cells.parquet        cell metadata + embeddings (loaded once at runtime)
  expr/<assay>/feature=*/…   feature-partitioned expression (queried per gene)
  manifest.yaml        the app's index of assays / embeddings / metadata
  config.yaml          optional defaults (edit to taste)
  app.R                deployable runtime: scroll_app(".")
```

Useful `scroll_build()` arguments:

* `assays = c("RNA", "ADT")` — export more than the default assay (multimodal /
  CITE-seq). Each becomes its own `expr/<assay>/` subtree and the app gains an
  **Assay** selector.
* `counts = TRUE` — also export a raw `counts/<assay>.parquet` store. This roughly
  doubles expression storage but enables the **Pseudobulk DE** panel (edgeR/
  limma-voom need counts).
* `meta_cols =` — choose which metadata columns to expose (default: a sensible set
  of categorical annotations + `nCount`/`nFeature`/`percent.*` QC columns).
* `quantize = FALSE` — keep full-precision expression (the default quantizes to
  `uint8` for compactness).

## 2. Customize the defaults

Every control auto-populates from the manifest; `config.yaml` only sets optional
defaults. Edit it after building:

```yaml
title: "PBMC 3k"          # app-bar label (omit for just the brand)
brand: "My Lab"           # replaces the "scroll" wordmark
default_embedding: umap
default_assay: RNA
markers: [CD3D, CD8A, MS4A1, CD14, NKG7]   # DotPlot's starting gene panel
```

## 3. Explore

Run it locally:

```{r serve}
scroll_serve("pbmc3k")                 # opens the app in your browser
```

The app is a scrolling page of analysis sections, each with its own controls:

* **DimPlot** — the embedding coloured by any metadata column.
* **FeaturePlot** — coloured by a gene's expression, or by a numeric metadata
  column, with colour-quantile clipping.
* **Biaxial** — pairwise scatters of numeric columns (hashtag / ADT / QC).
* **DotPlot / Violin / Proportions** — marker panels, distributions, composition.
* **DE** — live Wilcoxon markers for a contrast (table + volcano).
* **Pseudobulk DE** — edgeR/limma-voom (needs `counts = TRUE`).

Every categorical plot supports a **Manual** palette (per-level colour pickers),
every plot exports to PNG/PDF, and colours are assigned deterministically by level
name so a cell type keeps its colour across panels.

## 4. Subset views (optional)

If you have reprocessed a slice of the data in isolation — say the T cells,
re-embedded with their own UMAP and subclusters — expose it as a linked **View**.
`scroll_add_subset()` copies the child object's reduction (and any subset-only
metadata) onto the parent by barcode; `scroll_build(subsets = …)` records it.

```{r subsets}
obj <- scroll_add_subset(obj, tcell_obj, name = "tcell", label = "T cells",
                         embeddings = c(umap_tcell = "umap"),
                         meta = "tcell_subcluster")
scroll_build(obj, "pbmc3k")
```

Selecting *T cells* in the app bar restricts every panel to those cells and
switches the scatter panels to the sub-embedding.

## 5. Deploy to a Shiny Server

The built project **is** a deployable app — its `app.R` is just `scroll_app(".")`.
Copy the directory into your Shiny Server app root:

```{r deploy}
# on the server, after installing scroll + its runtime dependencies:
#   cp -r pbmc3k /srv/shiny-server/pbmc3k
# then browse to http://<server>:3838/pbmc3k/
```

The runtime does **not** need `SeuratObject` (only the build does), so a
viewer-only server stays lightweight. See the bundled `DEPLOY.md` in a built
project's parent for the full dependency list.

## 6. Query expression without the app

The same flat-RAM query layer is available programmatically:

```{r query}
con <- scroll_connect("pbmc3k")
hit <- scroll_query_feature(con, "RNA", "MS4A1")   # data.frame(cell, value)
man <- scroll_manifest("pbmc3k")
hit$value <- scroll_dequantize(hit$value, man, "RNA")
scroll_disconnect(con)
```

## Next

To add your own analysis section, see
`vignette("custom-panels", package = "scroll")`.
