---
title: "Large datasets & streaming builds"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Large datasets & streaming builds}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`scroll` is built for scale: the runtime app never loads the expression matrix, so
memory stays flat regardless of cell count. This vignette covers the two decisions that
matter once a dataset is large — **how to build it without running out of RAM**
(streaming), and **what to store** (the `quantize` / `counts` tradeoffs).

As a rough sense of scale, a multi-million-cell atlas builds on a 32 GB laptop and serves
in ~1.5 GB of runtime RAM, answering single-gene queries in tens of milliseconds.

## Why streaming

`scroll_build()` takes one in-memory Seurat object. That is fine up to a point, but
merging tens of samples into a single multi-million-cell object needs far more RAM than
the final store does — the merge, not the app, is the wall.

`scroll_build_stream()` sidesteps it: it reads **one source at a time**, appends that
source's expression to a shared store, and writes the manifest once at the end. Peak
memory is set by the *largest single source*, not the total — so a laptop that could
never hold the merged object builds the whole atlas comfortably.

## `scroll_build_stream()`

You supply the sources and a `reader` that turns each source into a (processed) Seurat
object; scroll streams them into `outdir`:

```{r stream-pattern}
library(scroll)

files <- Sys.glob("data/*.h5ad")            # e.g. one sample per file

scroll_build_stream(
  outdir    = "atlas",
  sources   = as.list(files),
  reader    = function(f) {
    # return a processed Seurat object: a log-normalized `data` layer,
    # the reductions, and the metadata columns you want to expose.
    seu <- schard::h5ad2seurat(f, use.raw = TRUE, load.obsm = TRUE)  # .raw = lognorm
    seu@reductions[["umap"]] <- seu@reductions[["Xumap_"]]           # tidy the name
    seu$sample <- tools::file_path_sans_ext(basename(f))
    seu
  },
  assays    = "RNA",
  embeddings = "umap",
  meta_cols = c("sample", "seurat_clusters", "percent.mt"),
  id_of     = function(f) basename(f))       # id used for append-tracking
```

Key properties:

- **Flat RAM.** One source is in memory at a time; the store is written incrementally.
- **Append-safe / idempotent.** Re-running adds only sources not already in the store, so
  you can build the first few samples, serve the app, and add the rest later with the same
  call — the app just reflects more cells.
- **Explicit columns.** Because the reader's objects may be heterogeneous, `assays` /
  `embeddings` / `meta_cols` are declared once (taken from the first source when `NULL`,
  except `meta_cols`, which is required).
- **Values are float32** (unquantized) — a global quantization ceiling is unknown while
  streaming, so `quantize = TRUE` is not supported here (see the tradeoffs below).

The reductions come from the reader's object. If your files share a **global** embedding
(e.g. an integrated UMAP computed across all samples), each source's coordinates already
live in the same space and the merged scatter is coherent.

## `quantize` and `counts`: what to store

Expression is stored sparsely and partitioned by feature, so a single-gene query reads
only that gene's data — never the whole matrix — and the store stays compact even at
several million cells. Projects built with an older `scroll` keep working unchanged; you
never have to rebuild. Two build options trade precision and features against disk:

| Option | Effect | Use when |
|---|---|---|
| `quantize = TRUE` (default in `scroll_build`) | values → `uint8`; ~4× smaller than float32, but floors values below ~`max/510` to zero | maximum compression; you don't need exact low-expression fractions |
| `quantize = FALSE` | values → `float32`, exact | fractions matter (DotPlot dot size, DE `pct`), or streaming (always float32) |
| `counts = TRUE` | also writes a raw-counts store (`counts/<assay>.parquet`) | you need replicate-aware **Pseudobulk DE** |

The counts store roughly doubles build size and needs a raw-counts layer in the object;
leave it off (`counts = FALSE`, the default) unless you use pseudobulk. Live Wilcoxon DE
works from the normalized store either way.

## Runtime memory at scale

The running app loads `cells.parquet` — metadata + embedding coordinates — into memory
once, and answers each expression query from disk. So the **only** thing that scales with
cell count at runtime is `cells.parquet`, at roughly **0.3 GB per million cells**;
everything else (the expression store) stays on disk and is read one partition per query.
Doubling the cell count adds a few hundred MB of metadata, not tens of GB of matrix — a
2–3M-cell explorer runs comfortably in ~1.5–2 GB.

Large scatters (millions of points) rasterize on screen automatically (via `scattermore`,
a Suggests) and still export as vector.

## A runnable example

Two small synthetic "sources" streamed into one store, then queried — the same shape as a
real multi-sample build, small enough to run here:

```{r runnable, eval = requireNamespace("SeuratObject", quietly = TRUE)}
library(scroll)
suppressPackageStartupMessages(library(SeuratObject))

# a tiny processed object: a log-normalized `data` layer + a umap + one meta column
mk <- function(n, tag) {
  set.seed(nchar(tag))
  cts <- Matrix::Matrix(matrix(rpois(30 * n, 2), nrow = 30,
           dimnames = list(paste0("g", 1:30), paste0(tag, "_c", seq_len(n)))),
           sparse = TRUE)
  o <- CreateSeuratObject(cts)
  o <- SetAssayData(o, layer = "data", new.data = log1p(cts))       # pretend lognorm
  emb <- matrix(rnorm(n * 2), ncol = 2,
                dimnames = list(colnames(o), c("UMAP_1", "UMAP_2")))
  o[["umap"]] <- CreateDimReducObject(emb, key = "UMAP_", assay = "RNA")
  o$batch <- tag
  o
}

dir <- tempfile("scroll-stream-")
scroll_build_stream(
  dir, sources = list(mk(150, "A"), mk(150, "B")),
  reader = function(s) s,                     # sources are already objects here
  assays = "RNA", embeddings = "umap", meta_cols = "batch",
  id_of = function(s) unique(s$batch), verbose = FALSE)

con <- scroll_connect(dir)
scroll_manifest(dir)$n_cells                  # 300 (both sources)
head(scroll_query_feature(con, "RNA", "g1"))  # one row per cell that expresses g1
scroll_disconnect(con)
```

## See also

- `vignette("getting-started", package = "scroll")` — the basic build → serve → deploy flow.
- `vignette("custom-panels", package = "scroll")` — adding your own analysis sections.
- `?scroll_build_stream`, `?scroll_build`, `?scroll_query_feature`.
