---
title: "Multimodal: VDJ, spatial & ATAC"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multimodal: VDJ, spatial & ATAC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`scroll`'s build → store → view layers are modality-agnostic, so a single-cell
project can carry more than gene expression. This vignette covers the modalities
beyond RNA: **VDJ** (paired single-cell TCR/BCR immune repertoire), **spatial**
(10x Visium / imaging-based), and **scATAC** (chromatin accessibility). Each is
opt-in at build time and auto-surfaces its panels only for projects that carry it.

## VDJ / immune repertoire

## The idea

VDJ is **per-cell metadata**: each cell already carries its clonotype id, V/J gene
segments, and CDR3 sequences as columns on the Seurat object. So ingestion needs no
new store type. At build time you describe those columns with `vdj_spec()` and pass
it to `scroll_build()`; the build **bakes a compact `repertoire/` Parquet store**
(the clone table plus pre-computed diversity, gene-usage residuals, and tissue
correlation) alongside the expression store, and records a `vdj` block in the
manifest. Four repertoire panels then **auto-appear** for that project — and only
for projects built with a `vdj` spec.

## Describe the repertoire with `vdj_spec()`

`vdj_spec()` maps your object's per-cell columns onto the schema the panels use.
Defaults follow common 10x naming and differ by `chain_type` (TCR vs BCR), so a
tidy 10x object needs only a minimal spec — override any column that differs.

```{r}
library(scroll)

spec <- vdj_spec(
  chain_type  = "TCR",              # TRBV/TRAV/TRBJ/TRAJ segment + label defaults
  group_col   = "celltype",         # summarise the repertoire within these groups
  clone_col   = "clonotype",        # per-cell clone id (else derived from the CDR3s)
  antigen_col = "antigen",          # optional: split rank-abundance / usage by antigen
  tissue_col  = "tissue",           # optional: a tissue split
  cluster_col = "seurat_clusters"   # optional: per-cluster diversity as well as per-group
)
```

Key arguments:

* **`chain_type`** — `"TCR"` (TRBV/TRAV/TRBJ/TRAJ) or `"BCR"` (IGHV/IGKV/IGLV…). It
  sets the default segment/CDR3 column names *and* the panel labels.
* **`group_col`** (required) — a categorical column (cell type / cluster) that the
  usage, diversity, and composition summaries are grouped by.
* **`clone_col`** — the per-cell clonotype id. If omitted, a clonotype is derived by
  pasting the CDR3 amino-acid columns.
* **`segments` / `cdr3`** — named `label = column` vectors of the V/J and CDR3
  columns. Both default by `chain_type`; pass your own if your object names them
  differently (e.g. `segments = c(TRBV = "v_call_vdj", TRBJ = "j_call_vdj")`).
* **`antigen_col` / `tissue_col` / `cluster_col` / `loc_col` / `broad_col`** —
  optional splits and annotations. `loc_col` + `broad_col` together enable the
  clone-frequency **tissue-correlation** view.
* **`carry`** — extra categorical columns to keep in the clone table so the
  CDR3-length panel can colour by them.

## Build

Pass the spec as `vdj =`:

```{r}
scroll_build(
  obj, "projects/tcell",
  assays     = "RNA",
  meta_cols  = c("celltype", "antigen", "tissue", "seurat_clusters"),
  vdj        = spec,
  overwrite  = TRUE
)
```

The build writes `projects/tcell/repertoire/` (`rep_cells.parquet`,
`diversity.parquet`, and — when derivable — `chisq.parquet` and
`tissue_corr.parquet`) and adds a `vdj:` block to `manifest.yaml`. Everything else
is a normal scroll build, so the eight RNA panels work as usual on the same project.

## What appears in the app

Serve as usual — the repertoire panels are added automatically:

```{r}
scroll_serve("projects/tcell")
```

Four panels surface below the RNA sections:

* **Clone overview** — clone-size **rank-abundance** curves and an
  **expansion-category** composition bar (Single / Small / Medium / Large /
  Hyperexpanded), split by antigen or group.
* **V/J gene usage** — per-group segment **frequency**, or the **chi-square
  standardized-residual** heatmap (which groups over/under-use which V/J genes).
  The segment picker lists the segments your `chain_type` defines.
* **CDR3 length** — CDR3 amino-acid **length distribution** (histogram or density),
  per chain and coloured by any categorical clone column.
* **Diversity** — Shannon / Simpson / clonality / Gini / top-clone-proportion per
  group (or per cluster if `cluster_col` was given), plus the tissue-correlation
  heatmap when `loc_col` + `broad_col` were supplied.

Because the panels are gated on the manifest's `vdj` block, an **RNA-only project
shows none of them** — the app surface adapts to the data.

## BCR

For B-cell receptor data, set `chain_type = "BCR"`; the defaults switch to
IGHV/IGKV/IGLV segment names and heavy/light CDR3 columns, and the panels relabel
accordingly. Override `segments`/`cdr3` if your object uses different column names.

## Notes

* The repertoire panels are **clone-level aggregates**, computed once at build time
  from the whole dataset; they are independent of the app-bar cell-subset filter
  (arbitrary cell subsets don't map onto pre-aggregated clone statistics).
* The chi-square residuals and diversity metrics are baked at build time, so the
  running app stays light — it only reads the small `repertoire/` tables.

# Spatial (10x Visium / imaging)

## The idea

Spot (or cell) coordinates are **just a 2-D embedding**, so once the tissue
coordinates are exported as a reduction, DimPlot and FeaturePlot already plot on
them. What spatial adds is the **tissue image**: at build time `scroll_build()`
extracts `GetTissueCoordinates()` into a `spatial` embedding — in image-pixel space,
y-oriented so the tissue is upright — and bakes the H&E array to a small raster
asset. The manifest records an `images` block and marks the embedding
`kind: spatial`, which surfaces a **Spatial** panel that draws spots over the tissue
image at a fixed aspect ratio.

## Build

`spatial_spec()` names the image to export (default: the first `Images(object)`),
so a standard Visium object needs only `spatial = spatial_spec()`:

```{r}
library(Seurat)
library(SeuratData)

brain <- LoadData("stxBrain", type = "anterior1")
brain <- NormalizeData(brain)               # scroll needs a normalized `data` layer

scroll_build(
  brain, "projects/brain",
  assays    = "Spatial",
  meta_cols = c("region", "nCount_Spatial", "nFeature_Spatial"),
  spatial   = spatial_spec(),               # extract coords + bake the tissue image
  overwrite = TRUE
)
```

The build writes `projects/brain/spatial/spatial.rds` (the tissue raster + its
pixel extent) and adds an `images:` block to `manifest.yaml`. Coordinates are
aligned to the baked image, so points overlay it directly.

Options:

* **`image`** — which image / FOV to export when the object has several (default:
  the first). Multi-section objects can be built per section.
* **`name`** — the embedding name (default `"spatial"`).
* **`image_asset = FALSE`** — export coordinates only (no tissue underlay); the
  Spatial panel then draws spots on a blank background.

## What appears in the app

```{r}
scroll_serve("projects/brain")
```

* **Spatial** panel — cells/spots in tissue space, coloured by a **gene** (type a gene)
  or any **metadata** column, with a spot-size slider and (when a tissue image was
  baked) a show/hide toggle, drawn with `coord_fixed()` so the tissue keeps its shape.
  **Drag a rectangle to zoom into a region; double-click to reset.**
* **DimPlot / FeaturePlot** also list the `spatial` reduction, so you get a tissue
  scatter through the standard panels too.

The Spatial panel is gated on the project having a spatial embedding, so non-spatial
projects never show it.

## Imaging platforms (Xenium / CosMx)

Imaging-based data carries **cell centroids** but usually no H&E raster. It works the
same way — `spatial_spec()` extracts the FOV coordinates into a `spatial` embedding —
and the Spatial panel surfaces without an image underlay (the toggle is hidden):

```{r}
xen <- LoadXenium("path/to/xenium/outs")
xen <- NormalizeData(xen)
scroll_build(xen, "projects/xenium", assays = "Xenium",
             meta_cols = "celltype", spatial = spatial_spec())
```

Since Xenium sections can hold hundreds of thousands of cells, the Spatial panel
rasterizes the point layer on screen (via `scattermore` when installed) and stays a
true vector plot on export.

## Notes

* The tissue raster is stored as a small `.rds` (dependency-free at build and
  runtime); the running app reads only that asset plus the normal expression store.
* Coordinate/image alignment (the lowres scale factor and y-orientation) is handled
  at build time against Seurat's accessors, so the runtime just plots points over
  the raster.

# scATAC (chromatin accessibility)

## The idea

A scATAC **peaks** assay is just an assay whose features are genomic peaks named
`chr-start-end`. Once it's exported, **FeaturePlot / DotPlot / Violin / DE work on
accessibility unchanged** (accessibility instead of expression). `atac_spec()` adds
the genomic context: it marks the assay `kind: peaks`, parses each peak's
chromosome/start/end from its name, and bakes a `peaks.parquet` annotation table —
with each peak's **nearest gene** when the assay is a Signac `ChromatinAssay` that
carries an annotation.

## Build

List the peaks assay in `assays=` (so its matrix is exported) and pass
`atac = atac_spec()`:

```{r}
library(Signac)

scroll_build(
  atac_obj, "projects/atac",
  assays    = "ATAC",              # the peaks (ChromatinAssay) matrix
  embeddings = "umap",             # an LSI-based UMAP
  meta_cols = "celltype",
  atac      = atac_spec("ATAC"),   # parse peak coords + nearest gene
  overwrite = TRUE
)
```

The build writes `projects/atac/atac/peaks.parquet` (peak → chr/start/end + nearest
gene) and marks the assay `kind: peaks` in the manifest. Nearest-gene requires the
`ChromatinAssay` to carry an annotation (e.g. via
`Signac::Annotation(obj) <- GetGRangesFromEnsDb(...)`); without one, the table is
coordinates-only and the panel falls back to region search.

## What appears in the app

```{r}
scroll_serve("projects/atac")
```

* **Peaks** panel — find a peak **By gene** (pick a gene → its nearby peaks, when
  nearest-gene was baked) or **By region** (a text filter over peak names, e.g.
  `chr1-8`), then colour the embedding by that peak's accessibility.
* **FeaturePlot** lists the peaks assay too, so you can type a peak name directly;
  **DotPlot / Violin / DE** compare accessibility across cell types with no extra work.

Gene-activity and motif matrices, if present as additional assays, ride the normal
multi-assay path. Coverage/genome-browser tracks and motif-PWM views are out of
scope.
